---
title: "TV Viewership"
author: "Travis Andersen"
output: pdf_document
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(forecast)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

<!-- There are numerous ways television networks measure their shows’ performances to determine whether they’re worthy of staying on the screen. One such metric is “overnight ratings” which refers to the amount of people who watch a show live, as it airs. For decades, TV networks have used these numbers as their guiding light. High overnight ratings are good, low overnight ratings are a death sentence. Since commercials are how shows primarily earn revenue, companies will often value this number above overall ratings. -->

<!-- The dataset Viewership.txt contains overnight ratings data for a popular TV show that I webscraped from the associated wikipedia page. The dataset contains overnight viewership rating for each of the shows n=70 episodes. The dataset contains the following variables: -->

<!-- Variable Name	Description -->
<!-- Season	The shows season -->
<!-- Episode	The episode number within the season -->
<!-- ShowNum	The overall show number -->
<!-- Viewers	The overnight viewers (in millions of people) -->

# 0.
<!-- Because the change in viewership is highly important in determining whether to keep producing the show, use the log-transformed Viewers variable in all your analysis below. This way, the change from one show to the next corresponds to a percentage increase or decrease in viewership. -->

```{r}
df <- read_table2("https://mheaton.byu.edu/docs/files/Stat469/Topics/2%20-%20TemporalCorrelation/1%20-%20TimeSeries/HWCaseStudy/Data/Viewership.txt") %>% 
  mutate(
    Viewers = log(Viewers)
  )
```

# 1.
<!-- Create exploratory plots and calculate summary statistics from the time series. Comment on any potential relationships you see between log(Viewers) and ShowNum (note, we are using ShowNum to denote “time” in this analysis). -->

```{r}
df %>% 
  ggplot(aes(x = ShowNum, y = Viewers)) +
  geom_line() +
  labs(
    x = "Show Number",
    y = "Log Viewers",
    title = "Scatterplot of show number against log number of viewers"
  )

df %>% 
  mutate(Episode = as.factor(Episode)) %>% 
  ggplot(aes(x = Episode, y = Viewers)) +
  geom_boxplot() +
  labs(
    x = "Episode",
    y = "Log Viewers",
    title = "Side-by-side boxplots of log viewers by episode"
  )

cor <- cor(df$ShowNum, df$Viewers)
```

From the line graph, we see that as the show goes on, log viewership increases. From the side-by-side boxplots, it seems that the episode number does not affect log viewership. The correlation between show number and log viewership is `r round(cor, 4)`, which is very high. 


# 2.
<!-- Fit a linear regression model to log(Viewers) using ShowNum as the explanatory variable. Determine if there is temporal correlation in the residuals which should be accounted for in your model. Discuss what this temporal correlation means for viewership. -->

```{r}
lm <- lm(Viewers ~ ShowNum, df)
acf <- acf(lm$residuals, plot = FALSE)
acf$acf %>% 
  as_tibble %>% 
  mutate(Lag = 0:18) %>% 
  rename(ACF = V1) %>% 
  ggplot(aes(x = Lag, y = ACF)) +
  geom_col() +
  ggtitle("Autocorrelation graph of model residuals")
```

To see if temporal correlation exists in the residuals, the autocorrelation in the residuals must be checked. From the graph above, we see that there is a significant amount of autocorrelation in the residuals. We can see that the amount of viewers who saw the last few episodes significantly affects the viewership of the current episode. 

# 3.
<!-- Fixing d=0 and D=1, determine appropriate values of p, q, P, Q in your time series model (note you should be able to figure out the seasonal cycle value S). Only consider p∈{0,1,2}, q∈{0,1,2}, P∈{0,1} and Q∈{0,1}. Discuss how you came to choose your specific values. -->

```{r}
ts <- ts(df$Viewers, c(1, 1), frequency = 10)
X <- model.matrix(Viewers ~ -1 + ShowNum, df)
best_model <- forecast::auto.arima(
  ts, d = 0, D = 1, max.p = 2, max.q = 2, max.P = 1, 
  max.Q = 1,  
  ic = "aic", stepwise = FALSE, xreg = X
)
```

To choose the appropriate order for the time series model, all SARIMA models with seasonality 10 and with $p \in {0, 1, 2}$, $q \in {0, 1, 2}$, $P \in {0, 1}$, and $Q \in {0, 1}$ were compared, and the model with the lowest aic was chosen. The SARIMA model which minimized aic was a SARIMA(2, 0, 0)(0, 1, 1) model. 

# 4.
<!-- Write down your selected time series regression model in terms of population parameters including your specification for the time series component of the residuals. Explain the meaning of any parameters in your model (including the time series components). Explain how statistical inference for your model can be used to predict the viewership moving forward. -->

The model can be written as follows:

$$\textbf{y} =  \textbf{X} \pmb{\beta} + \pmb{\epsilon}$$
$$\epsilon \sim SARIMA(2, 0, 0)(0, 1, 1)_{10} $$

$\pmb{\beta}$ is a vector containing the parameters $\beta_0$ and $\beta_1$. $\beta_0$ is the model parameter for the intercept and $\beta_1$ is the model parameter for the coefficient of the show number variable. $\epsilon$ follows a $SARIMA(2, 0, 0)(0, 1, 1)_{10}$ model. The model has a seasonality period of 10. It calculated the residuals based on the residuals of the last two episodes, and the unique part of the last episode, all after differencing the residuals by the residual of the same episode from the last season. We can predict the viewership moving forward by doing one-step ahead forecasts using the predicted residuals from the future, which will eventually converge to zero.  

# 5.
<!-- Fit your chosen time series model and validate any model assumptions you used. -->

```{r}
model <- Arima(ts, c(2, 0, 0), c(0, 1, 1), xreg = X)
df %>% 
  ggplot(aes(x = ShowNum, y = Viewers)) +
  geom_line() +
  geom_smooth(se = FALSE) +
  labs(
    x = "Show Number",
    y = "Log Viewers",
    title = "Scatterplot of show number against log number of viewers"
  )
```

From the line graph of the time series, we see that there is a fairly linear relationship.  

```{r}
acf_ts <- acf(model$residuals, plot = FALSE)
acf_ts$acf %>% 
  as_tibble %>% 
  mutate(Lag = 0:18) %>% 
  rename(ACF = V1) %>% 
  ggplot(aes(x = Lag, y = ACF)) +
  geom_col() +
  ggtitle("Autocorrelation graph of time series model residuals")
```

From an autocorrelation graph of the residuals from the time series model, we can see that there is no longer a significant correlation between the residuals. Therefore, the assumption of independence is met. 

```{r}
ggplot() + 
  geom_density(aes(x = model$residuals)) +
  labs(
    x = "Residuals",
    y = "Density",
    title = "Histogram of residuals"
  )
```

From the density plot above, we see that the residuals follow an approximately normal distribution.  

```{r}
ggplot() +
  geom_point(aes(x = model$fitted, y = model$residuals))
```

Disregarding the fitted values from the first season where they are fit exactly, the variance of the residuals does not depend on the fitted value amount, so equal variance is met.  

# 6.
<!-- Perform a cross-validation of predictions generated from your model for the most recent season of shows. Report the quality of your predictions in terms of RPMSE. Provide a plot of your predictions along with observed viewership and 95% prediction interval limits. -->

```{r}
n_cv <- 10
train <- df %>% 
  slice(1:(n() - n_cv))
test <- df %>% 
  slice((n() - n_cv + 1):n())


train_ts <- ts(train$Viewers, c(1, 1), frequency = 10)
train_X <- model.matrix(Viewers ~ -1 + ShowNum, train)
test_X <- model.matrix(Viewers ~ -1 + ShowNum, test)

train_model <- forecast::Arima(train_ts, order = c(2, 0, 0), seasonal = c(0, 1, 1), xreg = train_X)

fc <- forecast(train_model, h = n_cv, xreg = test_X, level = 0.95)

rpmse <- fc %>% 
  as_tibble %>% 
  bind_cols(test) %>% 
  summarize(
    rpmse = sqrt(mean((`Point Forecast` - Viewers)^2))
  ) %>% 
  pull(rpmse)

fc %>% 
  as_tibble %>% 
  mutate(ShowNum = 61:70) %>% 
  full_join(df, "ShowNum") %>% 
  ggplot(aes(x = ShowNum)) +
  geom_line(aes(y = Viewers)) +
  geom_ribbon(aes(ymin = `Lo 95`, ymax = `Hi 95`), fill = "blue", alpha = 0.2) +
  geom_line(aes(y = `Point Forecast`), col = "blue") +
  labs(
    x = "Show Number",
    y = "Viewers",
    title = "Data overlaid with predictions"
  )
```

A cross validation of the model over the last season of data was performed, and the calculated root predictive mean square error over the test period was `r round(rpmse, 4)`. 

# 7.
<!-- Determine if viewership is increasing or decreasing. Support your conclusions with appropriate hypothesis tests and confidence intervals. -->

```{r}
inf_df <- tibble(
  name = names(coef(model)),
  se = vcov(model) %>% 
    diag %>% 
    sqrt,
  mean = coef(model)
)

# p-value
p_value <- inf_df %>% 
  filter(name == "ShowNum") %>% 
  summarize(t = mean/se) %>% 
  pull(t) %>% 
  pt(df = nrow(df) - ncol(X) - 6, lower.tail = FALSE)

# confidence interval
ci <- inf_df %>% 
  filter(name == "ShowNum") %>% 
  summarize(
    lower = mean - se * qt(0.975, nrow(df) - ncol(X) - 6),
    upper = mean + se * qt(0.975, nrow(df) - ncol(X) - 6)
  )
```

A t-test with $H_0: \beta_{ShowNum} = 0$ and $H_A: \beta_{ShowNum} > 0$ returns a p-value of `r round(p_value, 4)`. A 95% confidence interval for $\beta_{ShowNum}$ is (`r round(ci$lower, 4)`, `r round(ci$upper, 4)`). Therefore, we conclude that viewership is increasing. 

# 8.
<!-- Season 8 is already in production. Forecast the log(Viewers) forward for season 8. Comment on how executives would be able to use these forecasts to gauge if the show should continue into a ninth season. -->

```{r}
future_X <- model.matrix(Viewers ~ -1 + ShowNum, tibble(ShowNum = 71:80, Viewers = 0))
future_fc <- forecast::forecast(model, xreg = future_X, level = 0.95)
future_fc %>% 
  as_tibble %>% 
  mutate(Episode = 1:10) %>% 
  knitr::kable(
    digits = 4,
    caption = "Season 8 viewership forecast"
  )
```

Executives can use these forecasts to gauge if the show should continue for a ninth season. These correspond to a percentage increase in viewership.  
